---
title: "Drug Mechanism Enrichment Analysis"
author: 
  - name: Belinda B. Garana
  - name: James H. Joly
package: DMEA
output:
  BiocStyle::html_document:
  vignette: >
    %\VignetteIndexEntry{Drug Mechanism Enrichment Analysis}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteIEncoding{UTF-8}
      toc_float: true
      toc_depth: 2
---

# Introduction
Though there are many algorithms for drug repurposing, most only evaluate individual drugs which can be prone to off-target effects. In contrast, our approach is to evaluate drug mechanisms-of-action by aggregating results across many drugs. Though mechanisms-of-action have also been evaluated in a similar manner by the Connectivity Map (clue.io), Drugmonizome (https://maayanlab.cloud/drugmonizome), and DrugEnrichr(https://maayanlab.cloud/DrugEnrichr), these tools only allow analysis of select public datasets and do not consider the ranks or directionality of input features. With our tool Drug Mechanism Enrichment Analysis (DMEA), users can query any dataset of interest using either an input gene signature or drug rank list to identify enriched mechanisms-of-action for drug repurposing efforts. For more information, please visit our website: https://belindabgarana.github.io/DMEA

# Installation
To install this package, start R and enter:
```{r eval=FALSE}
if (!require("BiocManager", quietlly = TRUE))
  install.packages("BiocManager")
  
BiocManager::install("DMEA")
```

Alternatively, to install this package directly from our GitHub repository, start R and enter:
```{r eval=FALSE}
if (!require("devtools", quietlly = TRUE))
  install.packages("devtools")
  
devtools::install_github("BelindaBGarana/DMEA")
```

If you are using Windows OS, you may need to change the above code to:
```{r eval=FALSE}
if (!require("devtools", quietlly = TRUE))
  install.packages("devtools")
  
devtools::install_github("BelindaBGarana/DMEA", build = FALSE)
```

# Input Option 1: Drug Rank List
With our drugSEA function, you can calculate the enrichment of drug sets in an input drug rank list. This drug rank list can be derived from anywhere (e.g., various drug repurposing algorithms). The only requirements are that:
  * Each drug has only one numeric, nonzero rank metric
  * At least 2 drug sets are represented in your list

Required inputs:
  * data: a dataframe containing one column for drug names and another column for drug ranks
  * gmt: gmt object containing drug set information. If not provided, a third column in your data input parameter must include set annotations; if so, the name of this column must be denoted by the set.type parameter. If you do not have your own drug set annotations, you can use our PRISM moa annotations on our GitHub repository at: https://raw.github.com/BelindaBGarana/DMEA/shiny-app/Inputs/MOA_gmt_file_n6_no_special_chars.gmt

Example:
```{r}
# prepare dataframe containing rank metrics and set annotations for each drug
Drug <- paste0("Drug_", seq(from = 1, to = 24))
# Drug is our default for the drug parameter

data <- as.data.frame(Drug)

data$Pearson.est <- rnorm(length(Drug), sd = 0.25)
# Pearson.est is our default for the rank.metric parameter

data$moa <- rep(paste("Set", LETTERS[seq(from = 1, to = 4)]), 6)
# moa is our default for the set.type parameter

# perform drugSEA and store results
# if you use other column names for your data parameter,
# then make sure to set drug, rank.metric, and set.type parameters
drugSEA.test <- DMEA::drugSEA(data)
```

# Input Option 2: Gene Signature
With our DMEA function, you can identify selectively toxic drug sets based on the correlations between drug sensitivity scores (e.g., AUC) and weighted gene voting scores generated with an input gene signature. This requires having drug screen and expression datasets which share the same samples (e.g., cell lines). To run this analysis, it is required that:
  * Each drug has only one numeric, nonzero correlation metric
  * At least 2 drug sets are represented in your drug screen data
  * Each gene has only one numeric weight value
  * At least one gene symbol matches between the input gene signature and expression data frames
  * At least 3 samples (e.g., cell lines) are in both the expression and drug screen sensitivity score data frames

Required inputs:
  * drug.sensitivity: a dataframe containing drug sensitivity scores (e.g., AUC) with drug names as column names, except for one column which contains sample names
  * gmt: gmt object containing drug set information. If not provided, a drug.info parameter with set membership information must be provided. If you do not have your own drug set annotations, you can use our PRISM moa annotations on our GitHub repository at: https://raw.github.com/BelindaBGarana/DMEA/shiny-app/Inputs/MOA_gmt_file_n6_no_special_chars.gmt
  * expression: a dataframe containing normalize gene expression with gene symbols as column names, except for one column which contains sample names
  * weights: a dataframe containing gene symbols in one column (default: column 1) and weight values in another column (default: column 2)

Example:
```{r}
# prepare drug sensitivity data frame
Sample_ID <- seq(from = 1, to = 21) # create list of sample names
drug.sensitivity <- as.data.frame(Sample_ID)
Drug <- paste0("Drug_", seq(from = 1, to = 24)) # create list of drug names
for(i in 1:length(Drug)){ # give each drug values representative of AUC sensitivity scores
  drug.sensitivity[,c(Drug[i])] <- rnorm(length(Sample_ID), mean = 0.83, sd = 0.166)
}

# prepare drug info data frame
# alternatively, a gmt object could be provided
info <- as.data.frame(Drug)
info$moa <- rep(paste("Set", LETTERS[seq(from = 1, to = 4)]), 6)
# moa is our default for the set.type parameter

# prepare gene weight data frame
Gene <- paste0("Gene_", seq(from = 1, to = 50)) # create list of gene symbols
weights <- as.data.frame(Gene) # by
weights$Rank_metric <- rnorm(length(Gene)) # give each gene a weight

# prepare expression data frame
# by default, column 1 of your expression data frame is the column name from which sample names are gathered
# the column containing sample names in your drug sensitivity data frame should have the same name
expr <- as.data.frame(Sample_ID)
for(i in 1:length(Gene)){ # give each gene values representative of normalized RNA expression
  expr[,c(Gene[i])] <- rnorm(length(Sample_ID), sd = 0.5)
}

# perform DMEA and store results
DMEA.test <- DMEA::DMEA(drug.sensitivity, expression = expr, weights = weights, drug.info = info)
```

# Session Info
```{r}
sessionInfo()
```
